---
title: "Daytime Observations"
format: 
  html:
    echo: false
editor: visual
---

```{r}
#| label: setup
#| message: false

library(sf)
library(tidyverse)
library(tmap)
sightings <- readRDS(here::here("data", "sightingsbins.rds"))
intervals <- readRDS(here::here("data", "intervals.rds"))
waypoints <- read_csv(here::here("data", "Waypoints.csv"))
waypoints_sf <- st_as_sf(waypoints, 
                         coords = c("Lon_dec", "Lat_dec"),
                         crs = "epsg:4326")
```

## **Species counts**

Excluding July 16, which revisited transect 15-16 and accounts for 75% of total sightings due to anomalous diving petrel aggregation. Only species with 15+ individuals shown here.

```{r}
#| label: species_counts

counts <- sightings %>% 
  group_by(species) %>% 
  summarize(count = sum(count))

filter(counts, count >= 15) %>% 
  mutate(species = fct_reorder(species, desc(count))) %>% 
  ggplot(aes(species, count)) +
  geom_col() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45),
        axis.title.x = element_blank())

```

## Species distributions

Maps of the 6 most numerous taxa: diving petrels (CODP+UNDP), fur seals, giant petrels (NGPT+SGPT+UGPT), cape petrels, South Georgia cormorants, and blue petrels. Also excludes July 16. Shown in 10 nm bins.

```{r}
#| label: species_maps
#| out-width: 100%
#| warning: false

recode_interval <- function(i, sz) floor((i - 1) / sz) + 1
interval_locs <- intervals %>%
  mutate(interval = recode_interval(interval, 10)) %>%
  group_by(transect, interval) %>%
  summarize(.groups = "drop") %>%
  st_centroid()
top6 <- sightings %>%
  as.data.frame() %>%
  mutate(interval = recode_interval(interval, 10),
         group = case_when(
           species %in% c("CODP", "UNDP") ~ "Diving petrel",
           species == "FUSE" ~ "Fur seal",
           species %in% c("UGPT", "NGPT", "SGPT") ~ "Giant petrel",
           species == "CAPT" ~ "Cape petrel",
           species == "SGCO" ~ "Cormorant",
           species == "BLPT" ~ "Blue petrel"
         )) %>%
  drop_na(group) %>%
  group_by(transect, interval, group) %>%
  summarize(count = sum(count, na.rm = TRUE),
            .groups = "drop") %>%
  right_join(interval_locs, by = c("transect", "interval")) %>%
  st_sf() %>% 
  drop_na(count)

sgmap <- terra::rast(here::here("data", 
                                "South Georgia Navigation Map", 
                                "SouthGeorgiaNavMap.tif"))
bbox <- st_bbox(c(xmin = -39.5, ymin = -55.5, xmax = -34.5, ymax = -52.5),
                crs = "epsg:4326")
for (g in unique(top6$group)) {
  p <- tm_shape(sgmap, bbox = bbox) +
    tm_rgb() +
    tm_shape(filter(top6, group == g)) +
    tm_symbols(size = "count",
               scale = 2,
               col = "#E41A1C",
               size.max = max(top6$count)) +
    tm_facets(by = "group") +
    tm_layout(g, legend.outside.position = "bottom")
  print(p)
}

```
