---
title: "Nighttime Observations"
format: 
  html:
    echo: false
    toc: true
editor: visual
---

```{r}
#| label: setup
#| message: false

library(readxl)
library(sf)
library(tidyverse)
library(tmap)
# Remove haul at waypoint 7x, it was aborted early due to weather
ikmt <- read_xlsx(here::here("data/PlanktonSamples.xlsx"),
                  sheet = "Stations") %>% 
  filter(Waypoint != "7x") %>% 
  mutate(Waypoint = round(parse_number(Waypoint), 2))
zoop <- read_xlsx(here::here("data/PlanktonSamples.xlsx"),
                  sheet = "Samples",
                  na = c("", "NA")) %>% 
  filter(Waypoint != "7x") %>% 
  mutate(Count = ifelse(Count > 9998, Inf, Count),
         Mass_g = parse_number(Mass_g),
         Waypoint = round(parse_number(Waypoint), 2))
# Estimate missing fish mass at waypoint by assuming Mass_g == Vol_mL
zoop$Mass_g[zoop$SampleID == 435] <- zoop$Volume_ml[zoop$SampleID == 435]
zoop$Mass_g[zoop$SampleID == 425] <- sum(zoop$Mass_g[zoop$SampleID %in% 426:435])
esuperba <- read_xlsx(here::here("data/PlanktonSamples.xlsx"),
                      sheet = "ESuperbaLengths") %>% 
  mutate(Waypoint = round(Waypoint, 2))
```

## Overall summary

We conducted `r nrow(ikmt)` IKMT net hauls, for a total of `r scales::comma(sum(zoop$Mass_g[!is.na(zoop$Phylum)], na.rm = TRUE) / 1000, accuracy = 0.1)` kg of plankton.

```{r}
#| label: summary
#| cap-location: top
#| tbl-cap: Frequency and total mass of major plankton groups in net hauls.

zoop %>% 
  filter(!is.na(Phylum)) %>% 
  mutate(group = case_when(
    Order == "Euphausiacea" ~ "Krill",
    Order == "Myctophiformes" ~ "Myctophids",
    Genus == "Themisto" ~ "Themisto",
    Class == "Cephalopoda" ~ "Squid",
    TRUE ~ "Other plankton"
  )) %>% 
  group_by(group) %>% 
  summarize(n_ikmt = n_distinct(Waypoint),
            mass_g = sum(Mass_g, na.rm = TRUE)) %>% 
  mutate(group = factor(group, levels = c("Krill", "Myctophids", "Squid", "Themisto", "Other plankton"))) %>% 
  arrange(group) %>% 
  set_names(c("Plankton", "Net hauls present", "Total mass (g)")) %>% 
  knitr::kable()

```

```{r}
#| label: group_zoop
zoop_grouped <- zoop %>% 
  filter(!is.na(Phylum)) %>% 
  drop_na(Mass_g) %>% 
  mutate(
    group = case_when(
      Order == "Euphausiacea" ~ "Krill",
      Order == "Amphipoda" ~ "Amphipods",
      Phylum == "Ctenophora" ~ "Ctenophores",
      Phylum == "Cnidaria" ~ "Cnidarians",
      TRUE ~ "Other"
    ),
    subgroup = case_when(
      Genus == "Thysanoessa" ~ "Thysanoessa",
      Species == "superba" ~ "E. superba",
      group == "Krill" ~ "Other krill",
      Genus == "Themisto" ~ "Themisto",
      group == "Amphipods" ~ "Other amphipods",
      TRUE ~ NA
    ),
    group = fct_reorder(group, Mass_g, sum, .desc = TRUE),
    subgroup = factor(subgroup, levels = c("Thysanoessa",
                                           "E. superba",
                                           "Other krill",
                                           "Themisto",
                                           "Other amphipods"))
  )

```

```{r}
#| label: zoop_coal

format_gen_sp <- function(genus, species) {
  str_glue("{substr(genus, 1, 1)}. {species}")
}

```

## Plankton biodiversity

```{r}
#| label: biodiversity

zoop_phylum <- zoop %>% 
  drop_na(Phylum) %>% 
  group_by(Phylum) %>% 
  summarize(n_ikmt = n_distinct(Waypoint)) %>% 
  mutate(Phylum = fct_reorder(Phylum, n_ikmt, .desc = TRUE))

ggplot(zoop_phylum, aes(Phylum, n_ikmt)) +
  geom_col() +
  labs(y = "Count IKMT") +
  theme_classic()

```

### Arthropod diversity

```{r}
#| label: arthro_diversity

arthro_diversity <- zoop %>%
  drop_na(Phylum) %>% 
  filter(Phylum == "Arthropoda") %>% 
  mutate(group = case_when(
    Genus == "Thysanoessa" ~ "Thysanoessa",
    Family == "Euphausiidae" ~ "Other krill",
    Genus == "Themisto" ~ "Themisto",
    Family == "Hyperiidae" ~ "Other hyperiids",
    Family == "Phrosinidae" ~ "Phrosinids",
    Family == "Lysianassidae" ~ "Lysianassids",
    Order == "Amphipoda" ~ "Other amphipods",
    Class == "Hexanauplia" ~ "Copepods",
    TRUE ~ "Other arthropods"
  )) %>% 
  group_by(group) %>% 
  summarize(n_ikmt = n_distinct(Waypoint)) %>% 
  mutate(group = fct_reorder(group, n_ikmt, .desc = TRUE))

ggplot(arthro_diversity, aes(group, n_ikmt)) +
  geom_col() +
  labs(y = "Count IKMT") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank())

```

### Fish diversity

```{r}
#| label: fish_diversity

is_egg <- function(label) {
  str_detect(label, "[Oo]rb") | str_detect(label, "[Ee]gg")
}

fish_diversity <- zoop %>%
  drop_na(Phylum) %>% 
  filter(Class == "Actinopterygii") %>% 
  mutate(group = case_when(
    Genus == "Electrona" ~ "Electrona",
    Genus == "Gymnoscopelus" ~ "Gymnoscopelus",
    Genus == "Protomyctophum" ~ "Protomyctophum",
    Genus == "Krefftichthys" ~ "Krefftichthys",
    Family == "Myctophidae" ~ "Other myctophids",
    Family == "Nototheniidae" ~ "Nototheniidae",
    is_egg(Sample) ~ "Egg",
    TRUE ~ "Other fish"
  )) %>% 
  group_by(group) %>% 
  summarize(n_ikmt = n_distinct(Waypoint)) %>% 
  mutate(group = fct_reorder(group, n_ikmt, .desc = TRUE))

ggplot(fish_diversity, aes(group, n_ikmt)) +
  geom_col() +
  labs(y = "Count IKMT") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank())

```

## Plankton biomass

Zooplankton biomass composition. The cnidarian biomass was mostly composed of two large individuals, an *Atolla* and a *Periphylla periphylla*, both found at waypoint 7.5.

```{r}
#| label: plankton_overall

zoop_grouped %>% 
  group_by(group) %>% 
  summarize(Mass_g = sum(Mass_g), .groups = "drop") %>% 
  ggplot(aes(group, Mass_g)) +
  geom_col() +
  labs(y = "Mass (g)") +
  theme_minimal() +
  theme(axis.title.x = element_blank())

```

Zooplankton biomass composition at each waypoint.

```{r}
#| label: plankton_by_waypoint

zoop_grouped %>% 
  mutate(
    Waypoint = fct_reorder(factor(Waypoint), Mass_g, sum, .desc = TRUE),
    group = fct_reorder(group, Mass_g, sum, .desc = TRUE)
  ) %>% 
  ggplot(aes(Waypoint, Mass_g, fill = group)) +
  geom_col() +
  scale_fill_brewer(palette = "Set2") +
  labs(y = "Mass (g)") +
  theme_minimal() +
  theme(legend.title = element_blank(),
        axis.text.x = element_text(angle = 90))

```

### Spatial distribution

```{r}
#| label: spatial

sgmap <- terra::rast(here::here("data", 
                                "South Georgia Navigation Map", 
                                "SouthGeorgiaNavMap.tif"))
bbox <- st_bbox(c(xmin = -39.5, ymin = -55.5, xmax = -34.5, ymax = -52.5),
                crs = "epsg:4326")

```

Map of total biomass in plankton tows.

```{r}
#| label: biomass_dist_total
#| out-width: 100%

total_biomass <- zoop %>% 
  filter(Sample == "Total") %>% 
  left_join(select(ikmt, Waypoint, Lat_dec, Lon_dec), by = "Waypoint") %>% 
  st_as_sf(coords = c("Lon_dec", "Lat_dec"), crs = "EPSG:4326")

tm_shape(sgmap, bbox = bbox) +
  tm_rgb() +
  tm_shape(total_biomass) +
  tm_symbols(size = "Mass_g",
             scale = 2,
             col = "goldenrod",
             border.col = "black",
             border.lwd = 1) +
  tm_layout(legend.position = c("right", "top"))

```

Krill biomass in plankton tows

```{r}
#| label: biomass_dist_krill
#| out-width: 100%

krill_biomass <- zoop %>% 
  filter(Family == "Euphausiidae") %>% 
  group_by(Waypoint) %>% 
  summarize(Mass_g = sum(Mass_g)) %>% 
  left_join(select(ikmt, Waypoint, Lat_dec, Lon_dec), by = "Waypoint") %>% 
  st_as_sf(coords = c("Lon_dec", "Lat_dec"), crs = "EPSG:4326")

tm_shape(sgmap, bbox = bbox) +
  tm_rgb() +
  tm_shape(krill_biomass) +
  tm_symbols(size = "Mass_g",
             scale = 2,
             col = "darkred",
             border.col = "white",
             border.lwd = 1) +
  tm_layout(legend.position = c("right", "top"))

```

## *Euphausia superba*

```{r}
#| label: superba_freq

superba_freq <- na.omit(zoop$Count[zoop$Species == "superba"])

```

*Euphausia superba* was found in 28 of the 43 net hauls. Number of individuals per haul was extremely skewed. 5 individuals or fewer were found in 22 hauls, and 20 or more in just 5 hauls. Overall, the mean and median number of individuals per haul were 13.2 and 2, respectively.

```{r}
#| label: superba_freq_plot

filter(zoop, Species == "superba") %>% 
  mutate(Waypoint = fct_reorder(as.character(Waypoint), 
                                Count, 
                                .desc = TRUE)) %>%
  ggplot(aes(Waypoint, Count)) +
  geom_col() +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90))

```

Mean *E. superba* length was 43.1 mm, ranging from \<30 mm to 60 mm. What does this say about age classes?

```{r}
#| label: superba_lengths

esuperba %>% 
  ggplot(aes(Length_mm)) +
  geom_histogram(binwidth = 1, fill = "grey80", color = "grey20") +
  geom_vline(xintercept = mean(esuperba$Length_mm), color = "red") + 
  theme_classic()

```

As would be expected, the mean length of *E. superba* in a net haul exhibited greater variance in hauls with fewer individuals. This variance is consistent with a normal distribution. In the figure below, each point represents the mean length of *E. superba* measured in a net haul and the error bars are the standard deviations. The blue lines indicate the 95% confidence interval of expected mean +/- standard deviation for a normal distribution with the same parameters as the observed distribution. The observed matches the expected well, suggesting the relatively small lengths measured for net hauls between 10 and 25 individuals is due to normal sampling error rather than measurement error.

```{r}
#| label: esuperba_var_n

esuperba_var_n <- esuperba %>% 
  group_by(Waypoint) %>% 
  summarize(n = n(),
            mean_length = mean(Length_mm),
            sd_length = ifelse(n > 2, sd(Length_mm), 0))

sims <- 10000
mean_length <- mean(esuperba$Length_mm)
sd_length <- sd(esuperba$Length_mm)
set.seed(1352)
expected <- tibble(
  n = rep(3:30, each = sims),
  sim_lengths = map(n, \(x) rnorm(x, mean_length, sd_length)),
  sim_mean = map_dbl(sim_lengths, mean),
  sim_sd = map_dbl(sim_lengths, sd)
) %>% 
  group_by(n) %>% 
  summarize(upr = quantile(sim_mean + sim_sd, 0.95),
            lwr = quantile(sim_mean - sim_sd, 0.05))

ggplot(esuperba_var_n, aes(n, mean_length)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), 
              expected, 
              fill = NA, 
              color = "blue") +
  geom_pointrange(aes(ymin = mean_length - sd_length, 
                      ymax = mean_length + sd_length),
                  position = position_jitter(width = 0.3)) +
  labs(x = "Individuals measured in net haul",
       y = "Mean length (mm)") +
  theme_classic()

```

The map of *E. superba* biomass shows far more biomass on the west side of South Georgia than the east.

```{r}
#| label: superba_map

superba_biomass <- zoop %>% 
  filter(Species == "superba") %>% 
  group_by(Waypoint) %>% 
  summarize(Mass_g = sum(Mass_g)) %>% 
  left_join(select(ikmt, Waypoint, Lat_dec, Lon_dec), by = "Waypoint") %>% 
  st_as_sf(coords = c("Lon_dec", "Lat_dec"), crs = "EPSG:4326")

tm_shape(sgmap, bbox = bbox) +
  tm_rgb() +
  tm_shape(superba_biomass) +
  tm_symbols(size = "Mass_g",
             scale = 2,
             col = "darkorange",
             border.col = "white",
             border.lwd = 1) +
  tm_layout(legend.position = c("right", "top"))

```
